\chapter{Fundamentals}
\label{ch:fundamentals}
The purpose of this chapter is to introduce, define, and describe the fundamental terms and concepts of this thesis proposal.
The terms and concepts defined within this chapter are assumed to be known in the following chapters. 
This chapter provides an introduction for access control in \autoref{sec:accesscontrol}.
Moreover, the relevant fundamentals of asymmetric cryptography are discussed in \autoref{sec:fundamentals:cryptography}.

\section{Access Control}
\label{sec:accesscontrol}
As stated by the \citeauthor{NIST2022} \cite{NIST2022}, access control is the process of granting and denying specific requests to logical or physical services and resources.
Based on the type of service or resource guarded by the access control, two types of access control can be distinguished.
Physical access control supervises access requests of subjects to specific physical facilities like federal buildings or military establishments.
Logical access control monitors and controls the access and usage of information and related information processing services.
Within the scope of the thesis proposal, the term access control is going to be used to describe logical access control for IT and OT systems.

In other words, as stated by \citeauthor{Hu2014} \cite{Hu2014}, logical access control protects objects like data, services, executable applications, or network devices from unauthorized operations.
An operation is performed by a subject on a specific object.
Operations include access, utilization, manipulation, and deletion of objects.
An operation may also be referred to as action.
To protect an object, the owners of the objects establish access control policies.
These policies describe which subjects may perform certain operations on a specific object.

The policies are enforced by logical components referred to as Access Control Mechanisms (ACM).
\citeauthor{Hu2014} \cite{Hu2014} state that the ACM receives the access request from the subject, decides whether the request should be granted or denied, and enforces the decision taken.
The ACM takes the decision based on a framework called access control model.
The access control model defines the functionalities and environment including subjects, objects, and rules for the ACM to take and enforce a decision.
In the following sections, five different access control models are introduced.
The access control models presented differ regarding their applicability and flexibility.
Moreover, each model has specific advantages and disadvantages.

\subsection{Attribute-Based Access Control (ABAC)}
The Attribute-Based Access Control (ABAC) is an access control model enabling access decisions based on attributes associated with subjects, objects, actions, and the environment of a system \cite{JTF2020}.
In other words, as stated by \citeauthor{Hu2014} \cite{Hu2014}, in ABAC an access request of a subject to perform operations on objects is decided based on assigned attributes of the subject and object, environment conditions, and a set of policies.
As a consequence, ABAC is also referred to as aspect-based access control \cite{Anderson2020} or policy-based access control.

Within the context of ABAC, an attribute is a characteristic containing information in the form of a name-value pair \cite{Hu2014}.
A subject attribute describes the characteristics of a person or non-person entity like identity, clearance, or department.
An object attribute describes the resource for which the access is requested, including the object classification, type, or owner.
An operation or action attribute describes the function performed on an object by a subject.
Operations include create, read, update, delete, or execute.
The environment conditions or environment attributes describe the context of an access request.
Environment conditions include dynamic characteristics like time of the day, day of the week, and request location of the subject.

A policy represents a rule based on which an access decision is taken for specific attributes \cite{Hu2014}.
As a consequence, a policy can be seen as a relationship between subject, object, environment, and operation attributes describing under which circumstances the ACM grants or denies an access request.

According to \citeauthor{Hu2014} \cite{Hu2014}, Role-Based Access Control (RBAC) and Identity-Based Access Control (IBAC) represent special cases of ABAC regarding their attributes used.
Furthermore, ABAC is capable of enforcing Discretionary Access Control (DAC) as well as Mandatory Access Control (MAC) concepts.
An advantage of ABAC compared to different access control models is the higher flexibility regarding multifactor policy expression.
Moreover, ABAC can take access control decisions based on ad-hoc knowledge and knowledge from separate infrastructure.
This is possible due to ABAC taking decisions at request time by evaluating policies instead of static decision-making as found in IBAC and RBAC.
As a consequence, pre-provisioning of requesting subjects in a multi-organization environment can be avoided.

\subsection{NIST Recommendations}
The National Institute of Standards and Technology (NIST) provides recommendations and guidance for authentication, authorization, and access control in ICS \cite{Stouffer2015} and OT systems \cite{Stouffer2023}.
With regard to authentication, NIST recommends considering the identity management lifecycle in OT environments.
This lifecycle includes the issuance, update, and revocation of authentication credentials.
Furthermore, NIST recommends the consideration of centralized identity management and authentication to improve management and monitoring.
NIST mentions Active Directory (AD) and Lightweight Directory Access Protocol (LDAP) as centralization supporting network technologies.
Nevertheless, NIST points out that authentication might not be advisable if it has an impact on performance, reliability, and safety.
This is especially the case in emergency situations in which authentication may impede procedures and result in negative consequences for system safety.

With regard to authorization and access control, NIST emphasizes the importance of considering physical and logical means for OT security.
NIST states that organizations are not limited to a single access control approach, but may rather employ different approaches resulting in higher effectiveness and efficiency.
A combination of ACLs, RBAC, and ABAC is mentioned as an example for achieving the access control requirements of an organization.

NIST recommends considering logical access control to minimize errors and costs of maintaining access privileges.
It is recommended that approaches support the Principle of Least Privilege (PoLP) and Separation of Duties (SoD).
Furthermore, NIST recommends solutions that incorporate credential management, authentication, authorization, access control, and system monitoring.
These solutions represent secure platforms enabling the access to OT devices.
Solutions that verify the identity of individuals or devices before granting access are recommended, as they lead to lower access and command processing latencies.
Moreover, solutions should be highly reliable and designed to reduce the impact on OT operations and safety.

\section{Asymmetric Cryptography (Public-Key Cryptography)}
\label{sec:fundamentals:cryptography}
Cryptography is a scientific discipline concerned with the study of methodologies, algorithms, schemes, and protocols for the encryption and verification of information~\cite{Barker2016,Barker2020,CNSS2022}.
In other words, cryptography provides means to prevent unauthorized access and to enable the verification of information.
The objective of cryptography is to satisfy specific security goals, including the assurance of confidentiality, integrity, authenticity, and non-repudiation.

A cryptographic system, also referred to as cryptosystem, is a set of cryptographic algorithms~\cite{Menezes1996}.
Moreover, a cryptosystem comprises sets of valid inputs and outputs as well as required cryptographic keys~\cite{Eckert2023}.
Two important principles for the design of cryptosystems were formulated by \citeauthor{Kerckhoffs1883} and \citeauthor{Shannon1949}.
As stated by \citeauthor{Kerckhoffs1883}, the cryptosystem must not require secrecy and must be able to be known by the adversary without inconvenience~\cite{Kerckhoffs1883}.
According to \citeauthor{Shannon1949}, it shall be assumed that the adversary knows the system being used~\cite{Shannon1949}.
The goal of a cryptosystem is to provide specific cryptographic services such as encryption or verification.
Verification describes the process of proving the integrity, authenticity, or non-repudiation of information~\cite{Boneh2023}.
The verification of information is based on a so-called tag or signature created by a signature algorithm.
Encryption describes the process of transforming plain information into an unintelligible form to maintain its secrecy~\cite{Barker2016,Boneh2023}.
The inverse process of encryption is referred to as decryption.
The intelligible plain information is referred to as plaintext.
The unintelligible or encrypted form of information is referred to as ciphertext.

Asymmetric cryptography, also referred to as Public-Key Cryptography (PKC), relies on algorithms which use a pair of two related keys for a cryptographic operation and its inverse operation~\cite{Barker2020,CNSS2022,Eckert2023}.
This pair of related keys consists of a private key and a public key.
The private key must be kept secret.
The public key may be shared without consequences for security, as long as its authenticity and integrity is ensured.
Although the two keys are related, the private key cannot be efficiently derived from the public key.

To encrypt a plaintext, a sending entity uses the public key of a receiving entity~\cite{Boneh2023}.
Only the receiving entity, whose public key was used for encryption, is able to decrypt the ciphertext using its own private key.
For the purpose of verification, so-called digital signatures are used in PKC.
Therefor, a sending entity computes a digital signature for a specific message using its own private key.
The digital signature is then appended to the message.
The receiver is able to prove the authenticity and integrity of the message by verifying the appended digital signature using the sender's public key.

PKC offers the following advantages over symmetric cryptography~\cite{Barker2020,Eckert2023}:
Firstly, PKC does not require a secure channel or secure protocol to exchange keys.
Secondly, the overall number of required keys using PKC is lower.
Moreover, the number of keys scales linear with the number of communication entities.
For example in a network with $n$ entities, $n$ key pairs or $2n$ keys have to be established.
In the same network, pairwise symmetric cryptography would require $n(n-1)/2$ keys.

Nevertheless, symmetric cryptography has advantages in comparison with asymmetric cryptography~\cite{Barker2020}.
Firstly, symmetric-key algorithms are faster than asymmetric-key algorithms.
Secondly, for a given level of security, symmetric cryptographic keys are shorter.
This reduces the memory and bandwidth requirements for key storage and transmission.

\subsubsection{Identity-Based Public-Key Cryptography}
The concept of identity-based cryptosystems and schemes was initially proposed by \textcite{Shamir1985}.
Identity-Based Public-Key Cryptography (ID-PKC) approaches allow the derivation of public keys from subject identities~\cite{Shamir1985,Boneh2023}.
Therefore, no certificates are required to verify that a public key belongs to a certain subject, since the subject identity is used to derive the public key.
In ID-PKC the existence of a public key does not depend on the existence of a corresponding private key.
In other words, the public key is not derived from the private key and may therefore exist before the private key.

ID-PKC employs a Trusted Third Party (TTP) called Private Key Generator (PKG) to generate private keys for entities~\cite{AlRiyami2003,Boneh2023}.
The PKG generates a system-wide key pair consisting of the master public key and master secret key.
The master public key is known to all entities.
The master secret key is only known to the PKG.
If an entity wishes to obtain its own private key, it has to prove its identity to the PKG.
If the identity is proven to the PKG successfully, the PKG generates a private key based on the entity's identity and the master secret key.

Since the PKG is able to generate private keys for arbitrary identities, ID-PKC leads to the key escrow problem~\cite{AlRiyami2003}.
This allows a misbehaving PKG to decrypt confidential messages or forge subject's signatures.
Key escrow is a mechanism that enables a TTP, e.g., a company, to retain the private component of a key pair~\cite{CNSS2021,Boneh2023}.
The goal of key escrow is to recover encrypted information, even if the encrypting entity and the corresponding secret key is not available.

\subsubsection{Certificateless Public-Key Cryptography}
Certificateless Public-Key Cryptography (CL-PKC) can be seen as an intermediate approach between ID-PKC and certificate-based PKC approaches such as Public Key Infrastructure (PKI)~\cite{AlRiyami2003}.
Certificate-based PKC approaches such as PKI use a TTP called Certificate Authority (CA) to issue, store, and revoke digital certificates~\cite{Eckert2023}.
These digital certificates are used to verify that a certain public key belongs to a specific subject which holds the corresponding private key.

CL-PKC approaches make use of a TTP called Key Generating Center (KGC) to generate partial private keys for entities based on the entity's identity and a master key~\cite{AlRiyami2003}.
To obtain the private key, the entity combines the partial private key with a secret value.
The generated private key is never shared with the KGC or other entities.
Consequently, CL-PKC neither suffers from the key escrow problem nor requires a secure communication channel for the key distribution.

To obtain the public key, the entity generates it based on public parameters and the secret value~\cite{AlRiyami2003}.
After the generation, the public key may be shared with other entities directly or via public directories.
Similar to ID-PKC, the public key is not derived from the private key and may therefore exist prior to it.
The only restriction is that the public key and the private key must use the same secret value.
However, CL-PKC is not identity-based, as the public key is not solely based on the identity of an entity.

\subsubsection{Attribute-Based Public-Key Cryptography}
Attribute-Based Public-Key Cryptography (AB-PKC) is a generalization of the ID-PKC concept~\cite{Sahai2005,Goyal2006,Hu2023}.
Attribute-Based Encryption (ABE) combines the principles of ABAC with the concept of PKC.
Therefor, attribute-based policies are integrated into cryptographic algorithms in the form of access structures and attributes.
This integration enables encryption based on arbitrary attributes and provides fine-grained access control via attribute-based decryption.
Communication complexity for key distribution in ID-PKC scales linear with the number of communication entities, i.e., users or devices.
In ABE the communication complexity scales linear with the number of attributes.
Similar to the concept of ABE, Attribute-Based Signatures (ABS) enable the integration of attributes into signing and verification algorithms~\cite{Li2010,Maji2011}.

ABE approaches are classified as either Key-Policy ABE (KP-ABE) or Ciphertext-Policy ABE (CP-ABE), depending on whether the access structure is associated with a key or a ciphertext~\cite{Goyal2006,Bethencourt2007,Hu2023}.
In KP-ABE the access structure is associated with a key.
The ciphertext is labeled with a set of attributes.
A user's secret key is able to decrypt the ciphertext if the attributes of the ciphertext satisfy the key-associated access structure.
Consequently, a data owner cannot control who is able to access the data and has to trust a TTP to issue appropriate keys~\cite{Bethencourt2007}.
In CP-ABE the access structure is associated with a ciphertext, and keys are associated with a set of attributes.
A user's secret key is able to decrypt the ciphertext if the key-associated attributes satisfy the ciphertext's access structure.
Accordingly, each data owner manages the access control policies for its own data, which makes CP-ABE more flexible and scalable than KP-ABE.
